Rice yield improvements through plant breeding are offset by inherent yield declines over time

### 2018-02-23 output/breed_paper_revised.Rnw
More revisions.

@@

Constraints on arable land are increasing simultaneous with the need
to increase total food production to meet a growing demand
\citep{Foley2011a, Godfray2010, Mueller2012}, which has necessitated
harvesting more grain per unit land area~\citep{Lobell2009,
Tittonell2014}. Historically, agricultural research has been
successful in staving off the ``Malthusian catastrophe'' of demand
surpassing supply via continued yield improvement of staple
crops. However, many production systems are experiencing plateaus in
grain yield~\citep{Grassini2013}. If this trend continues,
improvements per unit land area are no longer possible, and an
increase in the area under cultivation will be needed to meet food
demand, which carries with it undesirable ecological
implications~\citep{Foley2011a, tilman1999}. Therefore, it is of
critical importance to better understand why the rate of increase in
grain yields has declined or leveled off in intensified production
systems.

California (CA) rice represents one such production system. Rice is
grown primarily in the Sacramento Valley, which is characterized by
having a Mediterranean climate with long days, a dry growing season
relatively free of pests and diseases. These conditions lead to some
of the highest yields in the world~\citep{FAO2016}. Production in CA
is predominately focused on premium quality medium grain
\textit{japonica} cultivars (e.g., CalRose rice), and rice from CA is
recognized globally for its quality
(\url{http://agfax.com/2015/11/03/rice-calrose-wins-best-rice-world-competition/}). Most
cultivars in use in CA are developed by the California Cooperative
Rice Research Foundation (CCRRF; a collaboration between the
University of California, the USDA-Agricultural Research Service, and
farmer-funded research). In part due to improvements in rice genetics,
rice yields in CA increased rapidly during the period from 1920 --
1990; however, since the 1990s, the rate of yield increase has slowed
(Fig. \ref{fig:state_yield}) despite continuous crop
improvement. Based on a yield-gap analysis, on-farm yields in the
major rice growing region of CA is 73 to 76\% of maximum yield
potential~\citep{espe16a}. In highly intensive systems such as this,
\citet{grassini2011} have shown that farmers are capable of attaining
85\% of the maximum yield potential. Thus, given that farmers are near
the attainable yield potential, increases in yield are likely to be
slow and \citet{espe16a} reported that indeed this was the case
with yields on average increasing by about 50 to 62 kg ha\tsups{-1}
year\tsups{-1} between 1999 and 2014.

@@

The CCRRF Rice Experiment Station (RES) is located near Biggs, CA
(39.4648, -121.7342), and has been the central location for CCRRF's
efforts to develop improved rice cultivars adapted to CA since
1969. The California Statewide Variety Trails evaluate current and
promising cultivars at the RES and six to eight on-farm trials around
the state each year. To avoid potential complications due to site,
climate, and management differences at the farmer-managed trials
between years, we focused our investigation on the RES. The RES is
researcher-managed and plants the majority of the experimental plots
for the statewide program (approximately 3 to 4 times the number of
experimental plots compared to other on-farm locations). The climate
at the RES is Mediterranean, characterized by mild winters during
which most of the annual precipitation occurs (\Sexpr{18*24.5} to
\Sexpr{25*24.5} mm annually) and warm summers largely free of
precipitation events. Soils at the RES are classified as
Esquon-Neerdobe clays with roughly 2\% soil organic matter in the top
15cm~\citep{ssurgo}.

@@

>>>>

raw_samples = as.data.frame(RES_fullest)
@

@@

Constraints on arable land are increasing simultaneous with the need
to increase total food production to meet a growing demand
\citep{Foley2011a, Godfray2010, Mueller2012}, which has necessitated
harvesting more grain per unit land area~\citep{Lobell2009,
Tittonell2014}. Historically, agricultural research has been
successful in staving off the ``Malthusian catastrophe'' of demand
surpassing supply via continued yield improvement of staple
crops. However, many production systems are experiencing plateaus in
grain yield~\citep{Grassini2013}. If this trend continues,
improvements per unit land area are no longer possible, and an
increase in the area under cultivation will be needed to meet food
demand, which carries undesirable ecological
implications~\citep{Foley2011a, tilman1999}. Therefore, it is of
critical importance to better understand why the rate of increase in
grain yields has declined or leveled off in intensified production
systems.

California (CA) rice represents one such production system. Rice is
grown primarily in the Sacramento Valley, which is characterized by
having a Mediterranean climate with long days, a dry growing season
relatively free of pests and diseases. These conditions lead to some
of the highest yields in the world~\citep{FAO2016}. Production in CA
is predominately focused on premium quality medium grain
\textit{japonica} cultivars (e.g., CalRose rice), and rice from CA is
recognized globally for its quality
(\url{http://agfax.com/2015/11/03/rice-calrose-wins-best-rice-world-competition/}). Most
cultivars in use in CA are developed by the California Cooperative
Rice Research Foundation (CCRRF; a collaboration between the
University of California, the USDA-Agricultural Research Service, and
farmer-funded research). In part due to improvements in rice genetics,
rice yields in CA increased rapidly during the period from 1920 --
1990; however, since the 1990s, the rate of yield increase has slowed
(Fig. \ref{fig:state_yield}) despite continuous crop
improvement. Based on a yield-gap analysis, on-farm yields in the
major rice growing region of CA is 73 to 76\% of maximum yield
potential~\citep{espe16a}. In highly intensive systems such as this,
\citet{grassini2011} have shown that farmers are capable of attaining
85\% of the maximum yield potential. Thus, given that farmers are near
the attainable yield potential, increases in yield are likely to be
relatively slow and \citet{espe16a} reported that indeed this was the case
with yields on average increasing by about 50 to 62 kg ha\tsups{-1}
year\tsups{-1} between 1999 and 2014.

@@

The CCRRF Rice Experiment Station (RES) is located near Biggs, CA
(39.4648, -121.7342), and has been the central location for CCRRF's
efforts to develop improved rice cultivars adapted to CA since
1969. The California Statewide Variety Trials evaluate current and
promising cultivars at the RES and six to eight on-farm trials around
the state each year. To avoid potential complications due to site,
climate, and management differences at the farmer-managed trials
between years, we focused our investigation on the RES. The RES is
researcher-managed and plants the majority of the experimental plots
for the statewide program (approximately 3 to 4 times the number of
experimental plots compared to other on-farm locations). The climate
at the RES is Mediterranean, characterized by mild winters during
which most of the annual precipitation occurs (\Sexpr{18*24.5} to
\Sexpr{25*24.5} mm annually) and warm summers largely free of
precipitation events. Soils at the RES are classified as
Esquon-Neerdobe clays with roughly 2\% soil organic matter in the top
15cm~\citep{ssurgo}.

@@

\citet{mckenzie2014} explored the yield of rice in this same system
during a similar time period using similar data as these here, but
concluded that genetic rice yield potential has increased over this
time period at a rate of 25 kg\tsups{-1} ha\tsups{-1}
year\tsups{-1}. Yet the CI estimated from these data using the best
model suggests that there is only a
\Sexpr{round(sum((raw_samples$release_yr_c*scl)>=25)/nrow(raw_samples)*100,1)}\%
probability of the rate of yield increase is equal to or greater than
25 kg\tsups{-1} ha\tsups{-1} year\tsups{-1}. There are several
explanations for how this could be. 1) \citet{mckenzie2014} did not
account for yield decreases over time in their analysis. As shown
here, when these decreases are not taken into about, the estimated
increase in yield over time is estimated to be approximately the same
as that found by \citet{mckenzie2014}. 2) These yield decreases over time
were not taken into account possibly because \citet{mckenzie2014}
concluded that there was no trend in the yield of the major cultivars
over time. However this conclusion is based on a flawed interpretation
of the results of their statistical test. Failing to reject the
hypothesis that the trend in yield equals zero does not equate to the
confirmation of no trend. Instead, failing to reject the hypothesis
could suggest the analysis lacks the statistical power to detect the trend. 3) It
is not surprising that no trend in yield over time was detected given
that \citet{mckenzie2014} used data from several locations, yet
analyzed those data using a simple t-test. By using this simple
analysis, \citet{mckenzie2014} ignore systematic errors related to
differences between observations from different locations, including
different management systems, growing environments, and crop
timing. By contrast, this investigation presented here used a single,
well managed system and analyzed the data using a mixed effects model
which fully accounts for non-independence between similar
observations. In doing this, the analysis makes fewer assumptions and
more fully captures the structural relationships in the data. Hence,
the estimates presented here are likely more reliable.
